Standardization of ceramic shape: A case studay from the Iron Age pottery from northeastern Taiwan

### 2020-07-03 analysis/paper/paper.Rmd
edits on EFA methods

However, a key limitation of the landmark approach is that landmarks may be difficult to reproducibly locate for structures that are mostly or entirely curves if not mathematically-defined. In those cases, Elliptic Fourier Analysis (EFA) is an excellent mean for assessing morphological variation in whole structure of two-dimensional closed shapes. EFA has a long history in archaeology [@Gero1984], and has been increasingly used for the shapes of artifacts, such as stone artifacts and ceramics, and human remains and animal bones [@Fox2015; @Hoggard2019; @Iovictua2010; @Topi2018; @Wilczek2014]. EFA uses periodic functions to capture geometric information, where an outline is decomposed into a series of ellipses described by trigonometric functions [@Adams2004; @Bonhomme2014; @Claude2008]. To be specific, EFA converts coordinates along a curve into Fourier function coefficients, called harmonic coefficients or harmonics [@Kuhl1982]. The number of harmonics determines the quality and precision of the geometric representation of an object. The harmonic power, a cumulative sum of squared harmonic coefficients, provides a robust rule for determining the desired number of harmonics [@Bonhomme2014]. EFA is suitable for analyzing shapes lacking representative landmarks, or where curves contain the most meaningful variation. In such cases, the focus is the whole shape instead of the different components of a shape. We use EFA to evaluate the level of standardization of ceramics data from Kiwulan, northeastern Taiwan, in relation to the European presence in the 17th century to gain insights into the emergence of ceramic specialization. The reason of using EFA is because there is no sufficient landmarks could be identified straightforwardly due to the globular body of ceramics. We use a significance test for the equality of coefficient of variations of shape variables to statistically compare the vessel standardization from different periods.

>>>>

However, a key limitation of the landmark approach is that landmarks may be difficult to reproducibly locate for structures that are mostly or entirely curves if not mathematically-defined. In those cases, Elliptic Fourier Analysis (EFA) is an excellent mean for assessing morphological variation in whole structure of two-dimensional closed shapes. EFA was first introduced to study shapes of artifacts in archaeology in 1980s [@Gero1984]. Most applications are associated with human remains and animal bones to study biological variation, and has been increasing for stone artifacts to explore stone tool topology and technology [@Fox2015; @Hoggard2019; @Iovictua2010]. Despite the EFA applications for ceramics is rare, they demonstrate the potential of this promising approach for answering questions related to ceramic taxonomy [@Topi2018; @Wilczek2014]. For example, @Wilczek2014 evaluate the concordance between EFA and Discrete Cosine Transform (DCT), and a traditional typology by studying 154 complete ceramic vessels with varied shapes from the Bibracte oppidum in France. The results show the variation indicated by EFA and DCT matches the traditional ceramic typology, which supports that claim that outline-based approaches can be efficiently used for studying variations in ceramic shapes. Furthermore, @Wilczek2014's EFA results help us understand the level of production standardization over time across the region.

EFA uses periodic functions to capture geometric information, where an outline is decomposed into a series of ellipses described by trigonometric functions [@Adams2004; @Bonhomme2014; @Claude2008]. To be specific, EFA converts coordinates along a curve into Fourier function coefficients, called harmonic coefficients or harmonics [@Kuhl1982]. The number of harmonics determines the quality and precision of the geometric representation of an object. The harmonic power, a cumulative sum of squared harmonic coefficients, provides a robust rule for determining the desired number of harmonics [@Bonhomme2014]. EFA is suitable for analyzing shapes lacking representative landmarks, or where curves contain the most meaningful variation. In such cases, the focus is the whole shape instead of the different components of a shape. We use EFA to evaluate the level of standardization of ceramics data from Kiwulan, northeastern Taiwan, in relation to the European presence in the 17th century to gain insights into the emergence of ceramic specialization. The reason of using EFA is because there is no sufficient landmarks could be identified straightforwardly due to the globular body of ceramics. We use a significance test for the equality of coefficient of variations of shape variables to statistically compare the vessel standardization from different periods.
